Viscous Erosion in Porous Media

### 2018-03-25 docs/methods/erosion_methods.tex
edited some more

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Shear stress}
\label{sec:shearStress}
We have decomposed the deformation tensor into contributions from three
different terms.  The contributions from the Stokeslets and the
rotlets~\eqref{eqn:shearStressSR} can be immediately evaluated since
they do not involve any integration that needs to be approximated.  For
the shear stress of the double-layer potential, we use a slight variant
of the trapezoid rule. The kernel of the deformation tensor due to the
double-layer
potential~\eqref{eqn:shearStressDLP} is 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
This kernel has a singularity that is asymptotic to $\bigO(\rho^{-2})$
when $\yy \rightarrow \xx$, and this is too strong for the unmodified
trapezoid rule.  The singularity strength can be reduced since the
velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression has a singularity that is asymptotic to
$\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Then, to achieve
spectral accuracy, we apply the trapezoid rule with odd-even
integration~\cite{sid-isr1988}.  Therefore, the deformation tensor of
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_\ell$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_\ell}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_\ell \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  We also need to
approximate the jump in the deformation
tensor~\eqref{eqn:shearStressJump}, and this is done with Fourier
differentiation.  Finally, the shear stress is computed by multiplying
the deformation tensor by the normal vector, and taking the inner
product with the tangent vector.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Post processing quantities of interest}
\label{s:qoi}
\todo[inline]{Bryan is here}
To study characteristics of the flow, the vorticity
(Section~\ref{sec:vorticity}), pressure (Section~\ref{sec:pressure}),
and drag (Section~\ref{sec:drag}) will be computed.  Once
equation~\eqref{eqn:completed_DLP} is solved for the density function,
Stokeslets, and rotlets, all of these quantities can be computed.  For
target points in the fluid domain, the resulting layer potentials
involve integrands that are periodic and smooth, so the trapezoid rule
guarantees spectral accuracy~\cite{tre-wei2014}.  However, if the target
point is close to $\bd \Omega$, a nearly-singular integrand must be
integrated, and this requires upsampling that is described in
Section~\ref{sec:NSI}.  When the target points is on the boundary, the
integrands involve singularities that are evaluated by combining
singularity subtraction and odd-even integration~\cite{sid-isr1988}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vorticity}
\label{sec:vorticity}
We start by computing the vorticity $\omega(\xx) = v_x - u_y$ for $\xx \in
\Omega$.  As we did for the shear stress, we compute the vorticity due
to the double-layer potential, Stokeslets, and rotlets individually.
The vorticity of the double-layer potential~\eqref{eqn:velocityDLP} at
$\xx \in \Omega$ is
\begin{align}
  \omega^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} 
    \frac{(\rr \cdot \nn^\perp) + (\rr \cdot \nn)}{\rho^4}
    (\rr \cdot \eeta) ds_{\yy}.
  \label{eqn:vorticityDLP}
\end{align}
The vorticity due to the Stokeslet is
\begin{align*}
  \omega^S(\xx) = -\frac{1}{\pi} \sum_{\ell=1}^{M} 
    \frac{\rr \cdot \llambda_\ell^\perp}{\rho^2},
\end{align*}
and the vorticity of a rotlet is zero.  Therefore, the vorticity at $\xx
\in \Omega$ is
\begin{align*}
  \omega(\xx) = \omega^\DD(\xx) + \omega^S(\xx).
\end{align*}
To compute~\eqref{eqn:vorticityDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.

For the Stokes equations, on boundaries that have a no-slip boundary
condition, the shear stress is equivalent to the vorticity $\omega = v_x -
u_y$.  Therefore, we do not need to compute the vorticity for $\xx \in
\bd\Omega$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pressure}
\label{sec:pressure}
To compute the pressure at $\xx \in \Omega$, we follow the
same procedure used for the shear stress.  We start by computing the
pressure of the double-layer potential for $\xx \in \Omega$ and add the
appropriate jump if $\xx \in \Gamma$. For $\xx \in \Omega$, the pressure
is
\begin{align}
  p^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta(\yy) \,ds_\yy.
    \label{eqn:pressureDLP}
\end{align}
The pressure due to the Stokeslets is
\begin{align}
  p^S(\xx_0) = \sum_{\ell=1}^{M}\frac{\rr \cdot \llambda_\ell}{2\pi\rho^2},
  \label{eqn:shearPressureSR}
\end{align}
and the pressure due to the rotlets is zero.  Therefore, the pressure is
\begin{alignat*}{3}
  p(\xx) &= p^{\DD}(\xx) + p^{S}(\xx), &&\xx \in \Omega,\\
  p(\xx) &= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx), \quad && \xx \in \bd\Omega.
\end{alignat*}
To compute~\eqref{eqn:pressureDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.

To compute the pressure at $\xx_0 \in \bd\Omega$, we consider the limit
as $\xx$ tends to $\xx_0$.  The limiting value is given
by~\cite{qua-bir2014a}
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}p^\DD(\xx) =  
    \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0).
  \label{eqn:pressureJump}
\end{align}
Similar to the shear stress, the singularity of the pressure
corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is
too strong for the trapezoid rule to be directly applied.  Therefore, we
again subtract a constant density function and write the pressure as
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx)) \,ds_\yy, 
    \quad \xx \in \bd\Omega.
\end{align*}
Then, identical to how the shear stress is
computed~\eqref{eqn:stressOddEven}, odd-even integration is used to
evaluate this layer potential for pressure.  The pressure due to the
Stokeslets are given in equation~\eqref{eqn:shearPressureSR}.  Finally,
to add the jump condition~\eqref{eqn:pressureJump}, Fourier
differentiation is used to compute the arclength derivative.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Drag}
\label{sec:drag}
With the pressure and shear stress, the drag on a single body with
boundary $\gamma_\ell$ is
\begin{align}
\FFD = \int_{\gamma_\ell} p \nn + \tau \ss \,ds.
\label{eqn:drag}
\end{align}
The reader may notice an unusual sign convention for pressure in this formula, which is simply due to the normal direction pointing {\em into} the bodies.

@@

The layer potential~\eqref{eqn:genericLP} is approximated with the
trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than 5
arclength spacings from $\bd\Omega$. For points that are closer than 5
arclength spacings to $\bd\Omega$, we use an upsampled trapezoid rule.
The required upsampling of the geometry and the density function is done
with a Fourier interpolant.  To minimize the number of discretization
points that are used to approximate the layer
potential~\eqref{eqn:layerPotential}, we use the same upsampling rate
for all target points that are in the same dyadic interval
(Table~\ref{tbl:upsampling}).  For points that are closer than $5/16$ of
an arclength from $\bd\Omega$, a value of 0 is simply assigned to
$\phi$.
\newcolumntype{V}{>{\centering\arraybackslash}m{4em}}
\newcolumntype{N}{@{}m{0pt}@{}}
\begin{table}[htpb]
\centering
%\begin{tabular}{|c|ccccc|}
\begin{tabular}{|V|V|V|V|V|V|N}
  \hline
  $d(\xx,\bd\Omega)$ &
  $\left[5h,\infty\right]$ &
  $\left[\frac{5}{2}h,5h\right]$ &
  $\left[\frac{5}{4}h,\frac{5}{2}h\right]$ & 
  $\left[\frac{5}{8}h,\frac{5}{4}h\right]$ &
  $\left[\frac{5}{16}h,\frac{5}{8}h\right]$ & \\ [2ex] 
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 & \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate of our near-singular
integration scheme.  $d(\xx,\bd\Omega)$ is the distance between a point
$\xx \in \Omega$ and the boundary of the domain $\bd\Omega$, and $h$ is
an arclength spacing.}
\end{table}

In future work, upsampling will not be adequate.  For example, if
eroding bodies are also sedimenting, bodies will come too close to one
another for upsampling to be sufficient.  Therefore, we plan to
implement a near-singular integration scheme that uses either
near-singular interpolation~\cite{qua-bir2014a}, quadrature by
expansion~\cite{klo-bar-gre-one2013}, Barycentric
interpolation~\cite{bar-wu-vee2015}, or panel-based
quadrature~\cite{hel-oja2008a}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Shear stress}
\label{sec:shearStress}
We have decomposed the deformation tensor into contributions from three
different terms.  The contributions from the Stokeslets and the
rotlets~\eqref{eqn:shearStressSR} can be immediately evaluated since
they do not involve any integration to approximate.  For the shear
stress of the double-layer potential, we use a slight variant of the
trapezoid rule. The kernel of the deformation tensor due to the
double-layer potential~\eqref{eqn:shearStressDLP} is 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
This kernel has a singularity that is asymptotic to $\bigO(\rho^{-2})$
when $\yy \rightarrow \xx$, and this is too strong for the unmodified
trapezoid rule.  The singularity strength can be reduced since the
velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression has a singularity that is asymptotic to
$\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Then, to achieve
spectral accuracy, we apply the trapezoid rule with odd-even
integration~\cite{sid-isr1988}.  Therefore, the deformation tensor of
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_\ell$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_\ell}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_\ell \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  We also need to
approximate the jump in the deformation
tensor~\eqref{eqn:shearStressJump}, and this is done with Fourier
differentiation.  Finally, the shear stress is computed by multiplying
the deformation tensor by the normal vector, and taking the inner
product with the tangent vector.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Post processing quantities of interest}
\label{s:qoi}
We further characterize the flow by computing the vorticity
(Section~\ref{sec:vorticity}), pressure (Section~\ref{sec:pressure}),
and drag (Section~\ref{sec:drag}).  Once
equation~\eqref{eqn:completed_DLP} is solved, the vorticity, pressure,
and drag can be computed at arbitrary target points.  For target points
in the fluid domain, the integrands of the layer potentials are periodic
and smooth, and the trapezoid rule guarantees spectral
accuracy~\cite{tre-wei2014}.  However, if the target point is close to
$\bd \Omega$, a nearly-singular integrand must be integrated, and this
requires upsampling that is described in Section~\ref{sec:NSI}.  When
the target points is on the boundary, the integrands involve
singularities that are evaluated by combining singularity subtraction
and odd-even integration~\cite{sid-isr1988}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vorticity}
\label{sec:vorticity}
We start by computing the vorticity $\omega(\xx) = v_x - u_y$ for $\xx
\in \Omega$ which we decompose into contributions from the double-layer
potential, Stokeslets, and rotlets.  The vorticity of the double-layer
potential~\eqref{eqn:velocityDLP} at
$\xx \in \Omega$ is
\begin{align}
  \omega^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} 
    \frac{(\rr \cdot \nn^\perp) + (\rr \cdot \nn)}{\rho^4}
    (\rr \cdot \eeta) ds_{\yy}.
  \label{eqn:vorticityDLP}
\end{align}
To compute~\eqref{eqn:vorticityDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.
Next, the vorticity due to the Stokeslet is
\begin{align*}
  \omega^S(\xx) = -\frac{1}{\pi} \sum_{\ell=1}^{M} 
    \frac{\rr \cdot \llambda_\ell^\perp}{\rho^2},
\end{align*}
and the vorticity of a rotlet is zero.  Therefore, the vorticity at $\xx
\in \Omega$ is
\begin{align*}
  \omega(\xx) = \omega^\DD(\xx) + \omega^S(\xx).
\end{align*}

For the Stokes equations, on boundaries that have a no-slip boundary
condition, the shear stress is equivalent to the vorticity.  Therefore,
the vorticity is computed on the grains $\gamma_\ell$ by numerically
compute the shear stress as described in Section~\ref{sec:shearStress}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pressure}
\label{sec:pressure}
To compute the pressure at $\xx \in \Omega$, we follow the same
procedure used for the shear stress.  We first compute the pressure of
the double-layer potential for $\xx \in \Omega$ and add the appropriate
jump if $\xx \in \Gamma$. By taking the Laplacian of the Stokes
double-layer potential~\eqref{eqn:velocityDLP}, the pressure at $\xx \in
\Omega$ is
\begin{align}
  p^{\DD}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta(\yy) \,ds_\yy.
    \label{eqn:pressureDLP}
\end{align}
The pressure due to the Stokeslets is
\begin{align*}
  p^S(\xx_0) = \sum_{\ell=1}^{M}\frac{\rr \cdot \llambda_\ell}{2\pi\rho^2},
\end{align*}
and the pressure due to the rotlets is zero.  Therefore, the pressure is
\begin{align*}
  p(\xx) = p^{\DD}(\xx) + p^{S}(\xx), \quad \xx \in \Omega,
\end{align*}
To compute~\eqref{eqn:pressureDLP} numerically, we apply the trapezoid
rule with an upsampling scheme described in Section~\ref{sec:NSI}.

For target points on the boundary of the geometry, $\xx \in \bd \Omega$,
there is an additional term resulting from a jump in the pressure
double-layer potential~\cite{poz1992}
\begin{align}
  p(\xx)= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx), \quad \xx \in \bd\Omega.
  \label{eqn:pressureJump}
\end{align}
Similar to the shear stress, the singularity of the pressure
corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is
too strong when $\xx \in \bd\Omega$ for the trapezoid rule to be
directly applied.  Therefore, we again subtract a constant density
function and write the pressure as
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx)) \,ds_\yy, 
    \quad \xx \in \bd\Omega.
\end{align*}
Then, identical to how the shear stress is
computed~\eqref{eqn:stressOddEven}, odd-even integration is used to
evaluate this layer potential for pressure, and the jump condition
in~\eqref{eqn:pressureJump} is computed with Fourier differentiation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Drag}
\label{sec:drag}
Once the pressure and shear stress is computed, the drag on a single
body with boundary $\gamma_\ell$ is
\begin{align}
\FFD = \int_{\gamma_\ell} p \nn + \tau \ss \,ds.
\label{eqn:drag}
\end{align}
The reader may notice an unusual sign convention for pressure in this formula, which is simply due to the normal direction pointing {\em into} the bodies.

@@

The layer potential~\eqref{eqn:genericLP} is approximated with the
trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than five 
arclength spacings from $\bd\Omega$. For points that are closer than
five arclength spacings to $\bd\Omega$, we use an upsampled trapezoid
rule.  The required upsampling of the geometry and the density function
is done with a Fourier interpolant.  To minimize the number of
discretization points that are used to approximate the layer
potential~\eqref{eqn:layerPotential}, we use the same upsampling rate
for all target points that are in the same dyadic interval
(Table~\ref{tbl:upsampling}).  For points that are closer than $5/16$ of
an arclength from $\bd\Omega$, a value of 0 is simply assigned to
$\phi$.
\newcolumntype{V}{>{\centering\arraybackslash}m{4em}}
\newcolumntype{N}{@{}m{0pt}@{}}
\begin{table}[htpb]
\centering
%\begin{tabular}{|c|ccccc|}
\begin{tabular}{|V|V|V|V|V|V|N}
  \hline
  $d(\xx,\bd\Omega)$ &
  $\left[5h,\infty\right]$ &
  $\left[\frac{5}{2}h,5h\right]$ &
  $\left[\frac{5}{4}h,\frac{5}{2}h\right]$ & 
  $\left[\frac{5}{8}h,\frac{5}{4}h\right]$ &
  $\left[\frac{5}{16}h,\frac{5}{8}h\right]$ & \\ [2ex] 
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 & \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate of our near-singular
integration scheme.  $d(\xx,\bd\Omega)$ is the distance between a point
$\xx \in \Omega$ and the boundary of the domain $\bd\Omega$, and $h$ is
an arclength spacing.}
\end{table}

In future work, upsampling will not be adequate.  For example, if
eroding bodies are also sedimenting, bodies will come too close to one
another for upsampling to be sufficient.  Therefore, we plan to
implement one of several near-singular integration schemes available in
the literature~\cite{qua-bir2014a, klo-bar-gre-one2013, bar-wu-vee2015,
hel-oja2008a}.
